\bf Baseline Drift Estimation for Time Series Data Using Quantile Trend Filtering

### 2019-01-30 Manuscript/detrendify_JCGS.tex
manuscript updates

	\if0\blind
	{
		\title{\bf Drift Removal for Time Series Data Using Quantile Trend Filtering}
		\author{Halley Brantley\thanks{
				Department of Statistics, North Carolina State University, Raleigh, NC 27695 (E-mail: hlbrantl@ncsu.edu)} \,
			Joseph Guinness\thanks{
				Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, NY 14853 (E-mail: guinness@cornell.edu )} \,
			and
			%    and
			Eric C. Chi\thanks{Department of Statistics, North Carolina State University, Raleigh, NC 27695 (E-mail: eric$\_$chi@ncsu.edu).}    \\}
		\date{}
		\maketitle
	} \fi

@@

	\bigskip
	\begin{abstract}
		The text of your abstract.  200 or fewer words.
	\end{abstract}

	\noindent%
	{\it Keywords:}  3 to 6 keywords, that do not appear in the title
	\vfill

@@

	In many applications spanning the fields of chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar time series are observed and assumed to consist of a slowly varying trend other more rapidly varying components. Given observations $y(t)$, with $t=1, ..., N$, \cite{Kim2009} proposed using \textit{$\ell_1$ trend filtering} to estimate trends that are smooth in the sense of being piecewise linear or piecewise polynomial. \cite{Tib2014} then discovered that empirically the trend filtering estimates adapt to the local level of smoothness much better than the more common smoothing splines. In the trend filtering problem \citep{Kim2009, Tib2014}, the trend, $\theta \in \mathbb{R}^n$, is estimated by solving the following convex problem.
	\begin{eqnarray}
	\underset{\theta}{\arg\min}\; \frac{1}{2} \lVert y - \theta \rVert_2^2 + \lambda \lVert \mathbf{D}^{(k+1)}\theta \rVert_1,
	\end{eqnarray}
	where $\lambda \geq 0$ is a regularization parameter and the matrix $\mathbf{D}^{(k+1)} \in \mathbb{R}^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing $\mathbf{D}^{(k+1)}$ consider the difference operator when $k = 0$.
	\begin{eqnarray}
	\mathbf{D}^{(1)} = \begin{pmatrix}
	-1 & 1 & 0 & \cdots & 0 & 0 \\
	0 & -1 & 1 & \cdots & 0 & 0 \\
	\vdots & & & & & \\
	0 & 0 & 0 & \cdots & -1 & 1 \\
	\end{pmatrix}
	\end{eqnarray}
	Thus, $\lVert \mathbf{D}^{(1)}\theta \rVert_1 = \sum_{i=1}^{n-1} \lvert \theta_i - \theta_{i+1} \rvert$ which is just total variation denoising in one dimension. The penalty incentivizes solutions which are piece-wise constant. For $k \geq 1$, the difference operator $\mathbf{D}^{(k+1)} \in \mathbb{R}^{(n-k-1) \times n}$ is defined recursively as follows
	\begin{eqnarray}
	\mathbf{D}^{(k+1)} & = & \mathbf{D}^{(1)}\mathbf{D}^{(k)}.
	\end{eqnarray}
	By penalizing the $k+1$ fold composition of the discrete difference operator, we obtain solutions which are piecewise polynomials of order $k$. Trend filtering provides excellent estimates of trends when the only components of the time series are the trend and random noise (Fig. \ref{fig:trendfilter}). In some cases, time series may also include a rapidly varying signal component in addition to the trend and noise, in these cases the trend filtering estimate over-estimate the trend in the places where signal is present (Fig. \ref{fig:trendfilter}).

	\begin{figure}
		\centering
		\caption{Examples of trend filtering solutions. The true trend is shown in black, while the .estimated trend is shown in red.}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg1.png}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg2.png}
		\label{fig:trendfilter}
	\end{figure}

	One application in which the observed time series consists of a slowly varying trend, non-negative signal, and rapidly varying noise is the output of low cost air quality sensors. The use of low-cost, portable, air quality sensors has increased dramatically in the last decade. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours (Fig. \ref{fig:raw_spod}).

@@

	In this application, as well as those described in \cite{Ning2014}, \cite{marandi2015qualitative}, and \cite{pettersson2013algorithm}, it is not the mean trend that is desired but rather the baseline trend, which we can think of as the trend in a low quantile of the data.  \cite{Koenker1978} were the first to propose substituting the squared loss used in regression with the check loss function  (Eq. \ref{eq1}) to estimate a conditional quantile instead of the conditional mean.

	\begin{equation}
	\label{eq1}
	 \rho_{\tau}(z) = \sum_{i=1}^n z_i(1-\mathbf{I}(z_i<0)).
	\end{equation}

	A variety of approaches have been proposed for estimating quantile trends.  The regression vector $\theta$ is estimated by finding the minimizer of $\rho_{\tau}(y-X\theta)$ where is the check-loss function and $\mathbf{I}$ is the indicator function.   \cite{nychka1995nonparametric} estimated quantile splines by combining the smoothing spline penalty commonly used in non-parametric regression with the check-loss function used in qunatile regression:

	\begin{equation*}
	\rho_{\tau}(Y - f) + \lambda\int (f''(t))^2 dt, 
 	\end{equation*}

	where $Y(t)$ is a response variable observed at time $t$, $f(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degrees of smoothing.

	We propose to use the trend filtering penalty with the check loss function to produce a non-parametric quantile regression estimate for removing trends in time series. The formulation was proposed by \cite{Kim2009} as a possible extension of $\ell_1$-trend filtering but not studied. Moreover we extend the basic framework to ensure non-crossing while modeling multiple quantiles. We also implement a parallel ADMM algorithm for series that are too large to be computed simultaneously and proposed a modified criteria for choosing the smoothing parameter. We demonstrate through simulation studies that our proposed model provides better or comparable estimates of non-parametric quantile trends than existing methods and is a more effective method of drift removal for low-cost air quality sensors.

@@

	We combine the ideas of quantile regression and trend filtering, namely consider the case where the design $\mathbf{X}$ is the identity matrix. For a single quantile level $\tau$ the estimation of the quantile trend filtering model can be posed as the following optimization problem.
	\begin{eqnarray}
	\label{eq:quantile_trend}
	\underset{\theta}{\min}\; \rho_\tau(y - \theta) + \lambda \lVert \mathbf{D}^{(k+1)} \theta \rVert_1,
	\end{eqnarray}
	where $\lambda$ is a non-negative regularization parameter. We address the problem of choosing $\lambda$ in Section \ref{sec:lambda_choice}. As with the classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of free or commercial solvers. In many cases, including ours, we are interested in estimating multiple quantiles simultaneously. We also want to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$. Given quantiles $\{\tau_1, ..., \tau_J\}$ such that $\tau_1 < \tau_2 < ... < \tau_J$, the optimization problem becomes

	\begin{eqnarray}
	\label{eq:noncross_trend}
	\underset{\theta_1, ..., \theta_J}{\min}\; \sum_{j=1}^J \left [\rho_{\tau_j}(y - \theta_{j}) + 
	\lambda_j \lVert \mathbf{D}^{(k+1)} \theta_j \rVert_1 \right ] \\
	 \text{subject to: }\; \theta_{1i} \le \theta_{2i} \le ... \le \theta_{Ji} \text{ for all } i,
	\end{eqnarray}

	where $\theta_j \in \mathcal{R}^n$. The additional constraints are linear in the parameters so the non-crossing quantile trends can still be estimated by a number of available solvers. In the rest of this paper we rely on the commercial solver Gurobi \citep{gurobi}.

	The number of parameters to be estimated is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, all solvers will eventually break.

@@

	To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose an alternating direction method of multipliers (ADMM) algorithm for solving large problems in a piecewise fashion. The ADMM algorithm is fully described by \cite{boyd2011distributed, gabay1975dual, glowinski1975approximation}. We apply the consensus ADMM algorithm to the the quantile regression trend filtering problem given in Eq. \ref{eq:quantile_trend}, by dividing our observed series $y(t)$ with $t = \{1, ..., N\}$ into overlapping windows

	\begin{align*}
	\begin{cases}
	y_1(t) = y(t) & \mbox{if~~} 1 \le t \le u_{1}\\
	y_2(t) = y(t) & \mbox{if~~} l_{2} \le t \le u_{2} \\
	y_3(t) = y(t) & \mbox{if~~} l_{3} \le t \le u_{3} \\
	\cdots & \\
	y_M(t) = y(t) & \mbox{if~~} l_{M} \le t \le  N\\
	\end{cases}
	\end{align*}

	 with boundaries $1 < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < ...< N$. Given quantiles $\tau_1 < ... < \tau_J$ to be estimated we define $\theta_{j,m}(t)$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $M$ at time point $t$. In order to write out the constraint the that overlapping sections must be equal we define a consensus variable

	 \begin{align}
		 \bar{\theta}_{j,m} =&  g(\theta_{j, m-1}, \theta_{j,m}, \theta_{j,m+1}) \\
		 =& \begin{cases} 
			 \frac{\theta_{j,m-1}(t)+\theta_{j,m}(t)}{2} & \mbox{if~~} l_{m} \le t \le u_{m-1}  \\
			 \theta_{j,m}(t) & \mbox{if~~} u_{m-1} \le t \le l_{m+1}  \\
			 \frac{\theta_{j,m}(t)+\theta_{j,m+1}(t)}{2} & \mbox{if~~} l_{m+1} \le t \le u_{m}  
			 \end{cases},
	\end{align}
	defining $\theta_{j,M+1} = \theta_{j,M}$ and $\theta_{j,0} = \theta_{j,1}$. Our windowed quantile trend optimization problem can then be written as 
	 \begin{align}
		 \label{eq:quantile_windows}
		 &\sum_{m=1}^M\underset{\theta_{1,m}, ..., \theta_{J,m}}{\min}\; \sum_{j=1}^J \left [\rho_{\tau_j}(y_m - \theta_{j,m}) + 
		 \lambda_j \lVert \mathbf{D}^{(k+1)} \theta_{j,m} \rVert_1 \right ] \\
		 &\text{subject to: }\; \theta_{1,m}(t) \le \theta_{2,m}(t) \le ... \le \theta_{J,m}(t) \text{ for all } m,t \\
		 &\text{subject to: }\; \theta_{j,m}(t) = \bar{\theta}_{j,m}(t) \text{ for all } j, m, t
	 \end{align}
	 Given values for the Lagrange multiplier $\omega_{j,m}$ and the consensus variable $\bar{\theta}_{j,m}$ we can write the augmented Lagrangian for finding the trends in window $m$:
	 \begin{equation}
	 \mathcal{L}(\theta_{j,m}, \bar{\theta}_{j,m}, \omega_{j,m}) = \sum_{j=1}^J\rho_{\tau_j}(y_m - \theta_{j,m})+\lambda \lVert \mathbf{D}^{(k+1)}\theta_{j,m}\rVert_1 +  \omega_{j,m}^T(\theta_{j,m} - \bar{\theta}_{j,m}) + 
	 \frac{\gamma}{2}||\theta_{j,m} - \bar{\theta}_{j,m}||_2^2
	 \end{equation}

	 We then estimate the trend separately in each window while constraining the overlapping pieces of the trends to be equal according the algorithm 1.

	\begin{figure}[!h] 
		\centering
		\caption{Window boundaries and trends fit separately in each window.}
		\includegraphics[width = 0.8\linewidth]{Figures/overlapping_windows.png}
	\end{figure}

@@

	We used the stopping criteria described by \cite{boyd2011distributed}. The criteria are based on the primal and dual residuals which represent the residuals for the primal and dual feasibility, respectively. The primal and dual residuals are defined as

	\begin{align}
	&r_p^{(q)} = \sqrt{\sum_{m=1}^M\sum_{j=1}^J\lVert\theta_{j,m}^{(q)} - \bar{\theta}_{j,m}^{(q)}\rVert_2^2}\\
	&r_d^{(q)} = \gamma\sqrt{\sum_{m=1}^M \sum_{j=1}^J\lVert\bar{\theta}_{j,m}^{(q)} - \bar{\theta}_{j,m}^{(q-1)}\rVert_2^2}.
	\end{align}

	The primal residual, $r_p$, represents the difference between the trend values in the windows and the consensus trend value while the dual residual, $r_d$s represents the change in the consensus variable from one iterate to the next. The algorithm is stopped when

	\begin{align}
		&r_p^{(q)} < \epsilon_{abs}\sqrt{NJ} + \epsilon_{rel}\underset{m}{\max}\left[\max 
		\left(\sqrt{\sum_{j=1}^J \lVert\theta_{j,m}^{(q)}\rVert_2^2}, \sqrt{\sum_{j=1}^J \lVert \bar{\theta}_{j,m}^{(q)} \rVert_2^2} \right )\right]\\
		&r_d^{(q)} < \epsilon_{abs}\sqrt{NJ} + \epsilon_{rel}\sqrt{\sum_{m=1}^M\sum_{j=1}^J\lVert \omega_{j,m}^{(q)}\rVert_2^2}
	\end{align}

@@

	Timing experiments illustrate the advantages of using our ADMM algorithm even on datasets where solving the problem simultaneously is possible. We use from one to four windows for each data size with an overlap of 500. The windows algorithm was run until the stopping criteria were met using $\epsilon_{abs} = 0.01$ and $\epsilon_{rel} = 0.001$. For each data size, $n$, 25 datasets were simulated using the peaks simulation design described below and trends for three quantiles were fit simultaneously: 0.05, 0.1, and 0.15 using a $\lambda = n/5$.

@@

	\subsection{Regularization Parameter Choice}
	\label{sec:lambda_choice}
	Our method can easily handle missing data by changing the check loss function to output 0 for missing values. This allows us to leave out validation observations that can be used to select the tuning parameter $\lambda$ and to compare method performance on real data. A number of methods have been proposed for selecting the quantile regression smoothing spline tuning parameter \cite{yuan2006gacv}.  \cite{KoenkerNgPortnoy1994} relate $\lambda$ to the number of interpolated points $p_{\lambda} = \sum I(y_i = \widehat{g}_i(x_i))$, which can be thought of as active knots, they propose the Schwarz criterion for the selection of $\lambda$
	\begin{equation}
	\mbox{SIC}(p_{\lambda}) = \log[\frac{1}{n}\rho_{\tau}(y - \widehat{g}(x))] + \frac{1}{2n}p_{\lambda}\log n
	\end{equation}

	The traditional Bayesian Information Criterion (BIC) is given by 
	\begin{equation}
	\mbox{BIC}(s) = -2\log(L\{\hat{\theta}(s)\}) + \nu(s)\log n 
	\end{equation}	
	where $\theta(s)$ is the parameter $\theta$ with those components outside $s$ being set to 0, and $\nu(s)$ is the number of components in $s$. If we assume an asymmetric Laplace likelihood $L(y|\theta) = \left(\frac{\tau^n(1-\tau)}{\sigma}\right)^n\exp\left\{-\sum_i\rho_\tau(\frac{y_i - \theta_i}{\sigma})\right\}$ and the number of non-zero elements of $D\theta$ as $df$
	\begin{equation}
	\mbox{BIC}(df) = 2\frac{1}{\sigma}\rho_{\tau}(y-\theta) + \mbox{df}\log n
	\end{equation} 
	We can choose any $\sigma>0$ and have found empirically that $\sigma =  \frac{1-|1-2\tau|}{2}$ produces stable estimates. \cite{chen2008} proposed the extended BIC for large parameter spaces 
	\begin{equation}
	\mbox{BIC}_{\gamma}(s) = -2\log(L\{\hat{\theta}(s)\}) + \nu(s)\log n  + 2\gamma\log{P\choose j,}~~\gamma \in [0,1]
	\end{equation}
	where $P$ is the total number of possible parameters and $j$ is the number of parameters included in given model. We used this criteria with $\gamma = 1$, $P=n-k$ where $k$ is the order of the differencing matrix and $j = \nu(s)$ is the number of non-zero entries in $D^{(k)}\theta$.

	We used a single dataset to illustrate the difference between the scaled, unscaled and extended BIC criteria.

>>>>

	\if0\blind
	{
		\title{\bf Baseline Drift Estimation for Time Series Data Using Quantile Trend Filtering}
		\author{Halley Brantley\thanks{
				Department of Statistics, North Carolina State University, Raleigh, NC 27695 (E-mail: hlbrantl@ncsu.edu)} \,
			Joseph Guinness\thanks{
				Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, NY 14853 (E-mail: guinness@cornell.edu )} \,
			and
			%    and
			Eric C. Chi\thanks{Department of Statistics, North Carolina State University, Raleigh, NC 27695 (E-mail: eric$\_$chi@ncsu.edu).}    \\}
		\date{}
		\maketitle
	} \fi

@@

	\bigskip
	\begin{abstract}
		We address the problem of estimating smoothly varying baseline trends in time series data using quantile trend filtering. We extend the basic framework to ensure non-crossing while estimating multiple trends simultaneously and implement an alternating direction method of moments (ADMM) algorithm for series that cannot be processed simultaneously. Timing experiments demonstrate the usefulness of our ADMM algorithm even in cases where the series can be processed simultaneously. We also address smoothing parameter selection and propose a modified criteria based on the extended Bayesian Information Criteria. We demonstrate both through simulation studies and an application to low cost air quality sensor data that our model provides better or comparable estimates than existing methods of non-parametric quantile trends and improves signal classification of low-cost air quality sensor output. 
	\end{abstract}

	\noindent%
	{\it Keywords:} quantile regression, non-parametric, trend estimation, smoothing splines
	\vfill

@@

	In applications spanning the fields of chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar time series are observed and assumed to consist of a slowly varying trend and other more rapidly varying components. \cite{Kim2009} addressed the case when the series consisted of just a trend and a random noise component and proposed the use of \textit{$\ell_1$ trend filtering} to estimate trends that piecewise linear or piecewise polynomial. \cite{Tib2014} later demonstrated that empirically the trend filtering estimates adapt to the local level of smoothness better than the more common smoothing splines. In the trend filtering problem, given observations $y(t)$ with $t=1,...,N$, the trend, $\theta(t)$, is estimated by solving the following convex problem.
	\begin{eqnarray}
	\underset{\theta}{\arg\min}\; \frac{1}{2} \lVert y - \theta \rVert_2^2 + \lambda \lVert \mathbf{D}^{(k+1)}\theta \rVert_1,
	\end{eqnarray}
	where $\lambda \geq 0$ is a regularization parameter and the matrix $\mathbf{D}^{(k+1)} \in \mathbb{R}^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing $\mathbf{D}^{(k+1)}$ consider the difference operator when $k = 0$.
	\begin{eqnarray}
	\mathbf{D}^{(1)} = \begin{pmatrix}
	-1 & 1 & 0 & \cdots & 0 & 0 \\
	0 & -1 & 1 & \cdots & 0 & 0 \\
	\vdots & & & & & \\
	0 & 0 & 0 & \cdots & -1 & 1 \\
	\end{pmatrix}
	\end{eqnarray}
	Thus, $\lVert \mathbf{D}^{(1)}\theta \rVert_1 = \sum_{i=1}^{n-1} \lvert \theta_i - \theta_{i+1} \rvert$ which is just total variation denoising in one dimension. The penalty incentivizes solutions which are piece-wise constant. For $k \geq 1$, the difference operator $\mathbf{D}^{(k+1)} \in \mathbb{R}^{(n-k-1) \times n}$ is defined recursively as follows
	\begin{eqnarray}
	\mathbf{D}^{(k+1)} & = & \mathbf{D}^{(1)}\mathbf{D}^{(k)}.
	\end{eqnarray}
	By penalizing the $k+1$ fold composition of the discrete difference operator, we obtain solutions which are piecewise polynomials of order $k$.

	Trend filtering provides excellent estimates of trends when the only components of the time series are the trend and random noise (Fig. \ref{fig:trendfilter}). In some applications, however, time series may also include a rapidly varying signal component in addition to the trend and noise, in these cases the trend filtering estimate over-estimates the trend in the places where signal is present (Fig. \ref{fig:trendfilter}).

	\begin{figure}
		\centering
		\caption{Examples of trend filtering solutions. The true trend is shown in black, while the estimated trend is shown in red.}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg1.png}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg2.png}
		\label{fig:trendfilter}
	\end{figure}

	One application in which the observed time series consists of a slowly varying baseline, non-negative signal, and rapidly varying noise is the output of low cost air quality sensors. The use of low-cost, portable, air quality sensors has increased dramatically in the last decade. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Fig. \ref{fig:raw_spod} provides an example of 3 co-located SPod sensor outputs during a proximate release on a test range. While all of the sensors response to the pollutant signal, the amount of baseline drift varies from one node to another.

@@

	In this application, as well as those described in \cite{Ning2014}, \cite{marandi2015qualitative}, and \cite{pettersson2013algorithm}, it is not the mean trend that is desired but rather the baseline trend, which can be thought of as a trend in a low quantile of the data.  \cite{Koenker1978} were the first to propose substituting the squared loss typically used in regression with the check loss function  (Eq. \ref{eq:checkloss}) to estimate a conditional quantile instead of the conditional mean.

	\begin{equation}
	\label{eq:checkloss}
	 \rho_{\tau}(z) = \sum_{i=1}^n z_i(1-\mathbf{I}(z_i<0)).
	\end{equation}

	Later, \cite{KoenkerNgPortnoy1994} addressed the issue of estimated quantile curves by combining the check loss penalty with trend filtering penalty with $k = 2$ which produces quantile trends that are piecewise linear. Other approaches to estimating quantile trends have also been proposed. Rather than using the trend filtering $\ell_1$-norm to penalize the discrete differences, \cite{nychka1995nonparametric} used the smoothing spline penalty based on the $\ell_2$-norm solving the optimization problem in Eq. \ref{eq:smoothingspline}.

	\begin{equation*}
	\label{eq:smoothingspline}
	\rho_{\tau}(y - f) + \lambda\int (f''(t))^2 dt, 
 	\end{equation*}

	where $f(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degrees of smoothing. \cite{Oh2011} proposed an algorithm for solving this problem based on approximating the checkloss function with a similar differentiable function.

	\cite{Racine2017} take an entirely different approach to estimating quantile trends. They constrain the response to follow a smooth location scale model of the form $y(t) = a(t) + b(t)\epsilon_i$ and estimate the $\tau$\textsubscript{th} conditional quantile given $t = t_0$ using a kernel estimator and local linear approach.

	We propose to use the trend filtering penalty combined with the check loss function to produce quantile trends that are piecewise quadratic. The formulation was proposed by \cite{Kim2009} as a possible extension of $\ell_1$-trend filtering but not studied. Moreover we extend the basic framework to ensure non-crossing while modeling multiple quantiles. We also implement a parallel ADMM algorithm for series that are too large to be computed simultaneously and propose a modified criteria for choosing the smoothing parameter. We demonstrate through simulation studies that our proposed model provides better or comparable estimates of non-parametric quantile trends than existing methods and is a more effective method of drift removal for low-cost air quality sensors.

@@

	We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$ the estimation of the quantile trend filtering model can be posed as the following optimization problem.
	\begin{eqnarray}
	\label{eq:quantile_trend}
	\underset{\theta}{\min}\; \rho_\tau(y - \theta) + \lambda \lVert \mathbf{D}^{(k+1)} \theta \rVert_1,
	\end{eqnarray}
	where $\lambda$ is a non-negative regularization parameter. We address the problem of choosing $\lambda$ in Section \ref{sec:lambda_choice}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of free or commercial solvers. In many cases, including ours, we are interested in estimating multiple quantiles simultaneously. We also want to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$. Given quantiles $\{\tau_1, ..., \tau_J\}$ such that $\tau_1 < \tau_2 < ... < \tau_J$, the optimization problem becomes

	\begin{eqnarray}
	\label{eq:noncross_trend}
	\underset{\theta_1, ..., \theta_J}{\min}\; \sum_{j=1}^J \left [\rho_{\tau_j}(y - \theta_{j}) + 
	\lambda_j \lVert \mathbf{D}^{(k+1)} \theta_j \rVert_1 \right ] \\
	 \text{subject to: }\; \theta_{1}(t) \le \theta_{2}(t) \le ... \le \theta_{J}(t) \text{ for all } t,
	\end{eqnarray}

	where $\theta_j \in \mathcal{R}^n$. The additional constraints are linear in the parameters so the non-crossing quantile trends can still be estimated by a number of available solvers. In the rest of this paper we rely on the commercial solver Gurobi \citep{gurobi} and its R package implementation, but it can easily be substituted for a free solver such as the Rglpk package by \cite{rglpk}. The number of parameters to be estimated in this problem is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, all solvers will eventually break.

@@

	To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose an alternating direction method of multipliers (ADMM) algorithm for solving large problems in a piecewise fashion. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is fully described by \cite{boyd2011distributed}. We apply the consensus ADMM algorithm to the the quantile regression trend filtering problem given in Eq. \ref{eq:quantile_trend}, by dividing our observed series $y(t)$ with $t = \{1, ..., n\}$ into overlapping windows

	\begin{align*}
	y_1(t) = y(t) & \mbox{~~if~~} 1 \le t \le u_{1}\\
	y_2(t) = y(t) & \mbox{~~if~~} l_{2} \le t \le u_{2} \\
	\cdots & \\
	y_M(t) = y(t) & \mbox{~~if~~} l_{M} \le t \le  n\\
	\end{align*}

	 with boundaries $1 < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < ...< n$. An illustration is given in Fig. \ref{fig:windows}. Given quantiles $\tau_1 < ... < \tau_J$ to be estimated we define $\theta_{j,m}(t)$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $M$ at time point $t$. In order to write out the constraint that the overlapping sections must be equal we define a consensus variable

	 \begin{align*}
		 \bar{\theta}_{j,m} =&  g(\theta_{j, m-1}, \theta_{j,m}, \theta_{j,m+1}) \\
		 =& \begin{cases} 
			 \frac{\theta_{j,m-1}(t)+\theta_{j,m}(t)}{2} & \mbox{if~~} l_{m} \le t \le u_{m-1}  \\
			 \theta_{j,m}(t) & \mbox{if~~} u_{m-1} \le t \le l_{m+1}  \\
			 \frac{\theta_{j,m}(t)+\theta_{j,m+1}(t)}{2} & \mbox{if~~} l_{m+1} \le t \le u_{m}  
			 \end{cases},
	\end{align*}
	defining $\theta_{j,M+1} = \theta_{j,M}$ and $\theta_{j,0} = \theta_{j,1}$. Our windowed quantile trend optimization problem can then be written as 
	 \begin{align*}
		 \label{eq:quantile_windows}
		 &\sum_{m=1}^M\underset{\theta_{1,m}, ..., \theta_{J,m}}{\min}\; \sum_{j=1}^J \left [\rho_{\tau_j}(y_m - \theta_{j,m}) + 
		 \lambda_j \lVert \mathbf{D}^{(k+1)} \theta_{j,m} \rVert_1 \right ] \\
		 &\text{subject to: }\; \\
		 &\qquad \theta_{1,m}(t) \le \theta_{2,m}(t) \le ... \le \theta_{J,m}(t) \text{ for all } m,t \\
		 &\qquad \theta_{j,m}(t) = \bar{\theta}_{j,m}(t) \text{ for all } j, m, t
	 \end{align*}
	 Defining Lagrange multipliers $\omega_{j,m}$, and penalty parameter $\gamma > 0$ we can write the augmented Lagrangian for finding the trends in window $m$:
	 \begin{equation*}
	 \mathcal{L}(\theta_{j,m}, \bar{\theta}_{j,m}, \omega_{j,m}) = \sum_{j=1}^J\rho_{\tau_j}(y_m - \theta_{j,m})+\lambda \lVert \mathbf{D}^{(k+1)}\theta_{j,m}\rVert_1 +  \omega_{j,m}^T(\theta_{j,m} - \bar{\theta}_{j,m}) + 
	 \frac{\gamma}{2}||\theta_{j,m} - \bar{\theta}_{j,m}||_2^2
	 \end{equation*}
	 We then estimate the trend separately in each window, which can be done in parallel, while constraining the overlapping pieces of the trends to be equal as outlined in algorithm 1.

	\begin{figure}[!h] 
		\centering
		\caption{Window boundaries and trends fit separately in each window.}
		\includegraphics[width = 0.8\linewidth]{Figures/overlapping_windows.png}
		\label{fig:windows}
	\end{figure}

@@

	We measure convergence use the stopping criteria described by \cite{boyd2011distributed}. The criteria are based on the primal and dual residuals which represent the residuals for the primal and dual feasibility, respectively. The primal residual, 
	\begin{equation}
	r_p^{(q)} = \sqrt{\sum_{m=1}^M\sum_{j=1}^J\lVert\theta_{j,m}^{(q)} - \bar{\theta}_{j,m}^{(q)}\rVert_2^2},
	\end{equation}
	represents the difference between the trend values in the windows and the consensus trend value while the dual residual,
	\begin{equation*}
	r_d^{(q)} = \gamma\sqrt{\sum_{m=1}^M \sum_{j=1}^J\lVert\bar{\theta}_{j,m}^{(q)} - \bar{\theta}_{j,m}^{(q-1)}\rVert_2^2},
	\end{equation*}
	represents the change in the consensus variable from one iterate to the next. The algorithm is stopped when 
	\begin{align}
		&r_p^{(q)} < \epsilon_{abs}\sqrt{nJ} + \epsilon_{rel}\underset{m}{\max}\left[\max 
		\left(\sqrt{\sum_{j=1}^J \lVert\theta_{j,m}^{(q)}\rVert_2^2}, \sqrt{\sum_{j=1}^J \lVert \bar{\theta}_{j,m}^{(q)} \rVert_2^2} \right )\right]\\
		&r_d^{(q)} < \epsilon_{abs}\sqrt{nJ} + \epsilon_{rel}\sqrt{\sum_{m=1}^M\sum_{j=1}^J\lVert \omega_{j,m}^{(q)}\rVert_2^2}
	\end{align}

@@

	Timing experiments illustrate the advantages of using our ADMM algorithm even on datasets where solving the problem simultaneously is possible. For each data size, $n$, 25 datasets were simulated using the peaks simulation design described below and trends for three quantiles were fit simultaneously: 0.05, 0.1, and 0.15 using a $\lambda = n/5$. We use from one to four windows for each data size with an overlap of 500. The windows algorithm was run until the stopping criteria were met using $\epsilon_{abs} = 0.01$ and $\epsilon_{rel} = 0.001$.

@@

	\subsection{Regularization Parameter Choice}
	\label{sec:lambda_choice}
	An important problem in trend estimation is the choice of regularization parameter or degree of smoothness. Our method can easily handle missing data by defining the check loss function to output 0 for missing values. This allows us to leave out validation observations that can be used to select the tuning parameter $\lambda$. However, the use of an information criteria metric can result in a better choice of regularization parameter than the validation method.  \cite{KoenkerNgPortnoy1994} addressed the choice of regularization parameter by proposing the Schwarz criterion for the selection of $\lambda$
	\begin{equation}
	\label{eq:SIC}
	\mbox{SIC}(p_{\lambda}) = \log\left[\frac{1}{n}\rho_{\tau}(y - \theta)\right] + \frac{1}{2n}p_{\lambda}\log n.
	\end{equation}
	where $p_{\lambda} = \sum_t I(y(t) = \widehat{\theta}(t))$ is the number of interpolated points, which can be thought of as active knots. The SIC is based on the traditional Bayesian Information Criterion (BIC) which is given by 
	\begin{equation}
	\mbox{BIC}(s) = -2\log(L\{\hat{\theta}\}) + \nu\log n 
	\end{equation}	
	where $L$ is the likelihood function and $\nu$ is the number of non-zero components in $\hat{\theta}$. If we take the approach used in Bayesian quantile regression \citep{yu2001bayesian}, and assume that minimizing the checkloss function corresponds to maximizing the asymmetric Laplace likelihood, \begin{equation}
	L(y|\theta) = \left(\frac{\tau^n(1-\tau)}{\sigma}\right)^n\exp\left\{-\sum_t\rho_\tau\left(\frac{y(t) - \theta(t)}{\sigma}\right)\right\},
	\end{equation} 
	we can compute the BIC as 
	\begin{equation}
	\mbox{BIC}(df) = 2\frac{1}{\sigma}\rho_{\tau}(y-\hat{\theta}) + \mbox{df}\log n
	\end{equation} 
	where $df$ is the number of non-zero elements of $D^{(k+1)}\hat{\theta}$. We can choose any $\sigma>0$ and have found empirically that $\sigma =  \frac{1-|1-2\tau|}{2}$ produces stable estimates.
